\pkg{Rgbp}: An R Package for Gaussian, Poisson, and Binomial Hierarchical Modeling

### 2013-11-20 jss/article.tex
code
edits

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Joseph Kelly\\
  Department of Statistics\\
  Harvard University\\
  1 Oxford Street, Cambridge, MA\\
  E-mail: \email{kelly2@fas.harvard.edu}\\
  URL: \url{http://www.people.fas.harvard.edu/~kelly2/}\\

@@

where $\mu_{0j} = x_{j}' \mathbf{\beta}_{(m\times1)}=\beta_{0}+\beta_{1}x_{1, j} + \cdots + \beta_{m-1}x_{m-1, j}$, $j=1, \ldots k$ and $m$ is the number of regression coefficients including an intercept term. The notation $x_{j}'=(1, x_{1, j}, x_{2, j}, \ldots, x_{m-1, j})$ represents the $j$-th group's $m\times 1$ covariate vector whose first element is 1 for the intercept term of regression coefficients. It is assumed that the second-level variance $A$ is unknown and that $\mathbf{\beta}_{(m\times1)}$ is also unknown unless specified otherwise. Note that if there are no covariates then $x_{j}=1$, ($m=1$) and so $\mu_{0j}=\mu_{0}=\beta_{0}$ for all $j$, resulting in an exchangeable prior distribution. For reference, a parameter with a zero subscript, such as $\mu_{0j}$, represents a mean parameter of the prior (second-level) distribution, \emph{i.e.}, a prior mean. Based on this conjugate prior distribution it is easy to derive the corresponding posterior distribution

\begin{equation} \label{normalpost}
\mu_{j}\vert \textbf{y}, \beta, A \stackrel{ind}{\sim}\textrm{Normal}(~(1-B_{j})y_{j} + B_{j}\mu_{0j},~(1-B_{j})V_{j}~),
\end{equation}
where $B_{j}\equiv V_{j}/(V_{j} + A),~j=1, \ldots, k$, are called shrinkages. By integrating over $\mu_{j}$ for $j = 1 \ldots k$ a marginal distribution, useful for model checking (Section \ref{discussion}), can be derived
\begin{equation} \label{normalmarginal}
y_{j}\vert \beta, A \stackrel{ind}{\sim}\textrm{Normal}(\mu_{0j}, V_{j} + A). 
\end{equation}

\subsection[Poisson-Gamma]{Poisson-Gamma (``p''=Poisson data)}\label{poissonsubsec}
\code{gbp} is also able to build a Poisson-Gamma hierarchical model, though its usage is limited to the case where the second-level mean is known ($m=0$). For notational consistency, let's define $y_{j}\equiv z_{j} / n_{j}$ as the unbiased estimate of $\lambda_{j}$ for the $j$-th group, where  $z_{j}$ is the number of successes (or failures) out of $n_{j}$ trials. For this Poisson model $n_{j}$ can be interpreted as the exposure of the $j$-th group,  not necessarily an interger. The notation $r$ represents the amount of prior information as $n_{j}$ does the amount of observed information. Note that the constant, $1/r$, multiplied to the Gamma distribution below is a scaling factor and the square brackets indicates the [mean, variance] of a distribution. 
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}\label{gammaprior}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(r\lambda_{0})\sim \textrm{Gamma} \left[\lambda_{0}, ~\frac{\lambda_{0}}{r} \right],
\end{equation}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 3,   2,   5,  11,   9,  12,  12,   4,  10,  13,  14,   7,  12,
          11,  13,  22,  15,  11,  14,  11,  16,  14,   9,  15,  13,  35,
          26,  25,  20,  35,  27)
R> n <- c(67,  68, 210, 256, 269, 274, 278, 295, 347, 349, 358, 396, 431,
         441, 477, 484, 494, 501, 505, 540, 563, 593, 602, 629, 636, 729,
         849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}
or
\begin{CodeChunk}
\begin{CodeInput}
R> data(hospitals)
R> z <- hospitals$d # variable name for the number of deaths is d in dataset
R> n <- hospitals$n
\end{CodeInput}
\end{CodeChunk}

In addition, suppose one knows that the state-level death rate per exposure of this surgery is 0.030 ($\lambda_{0}$). Using these data and the known second-level mean ($\lambda_{0}$), \pkg{Rgbp} provides point and interval estimates of the true death rate ($\lambda_{j}$) for each hospital so that one can evaluate the hospital reliabilities.

@@

\subsection[Unknown Second-level Mean and No Covariate]{8 Schools: Unknown Second-level Mean and No Covariate} \label{sec:ex:8schools}

@@

Due to it's speed and ease of use, \pkg{Rgbp} can be used for as a method of preliminary data analysis. Such results may tell statisticians and practitioners alike whether a more intensive method in terms of implementation and computational time, such as MCMC, is needed.

>>>>

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Joseph Kelly\\
  Department of Statistics\\
  Harvard University\\
  1 Oxford Street, Cambridge, MA\\
  E-mail: \email{kelly2@fas.harvard.edu}\\

@@

where $\mu_{0j} = x_{j}' \mathbf{\beta}_{(m\times1)}=\beta_{0}+\beta_{1}x_{1, j} + \cdots + \beta_{m-1}x_{m-1, j}$, $j=1, \ldots k$ and $m$ is the number of regression coefficients including an intercept term. The notation $x_{j}'=(1, x_{1, j}, x_{2, j}, \ldots, x_{m-1, j})$ represents the $j$-th group's $m\times 1$ covariate vector whose first element is 1 for the intercept term of regression coefficients. It is assumed that the second-level variance $A$ is unknown and that $\mathbf{\beta}_{(m\times1)}$ is also unknown unless otherwise specified. Note that if there are no covariates then $x_{j}=1$, ($m=1$) and so $\mu_{0j}=\mu_{0}=\beta_{0}$ for all $j$, resulting in an exchangeable prior distribution. For reference, a parameter with a zero subscript, such as $\mu_{0j}$, represents a mean parameter of the prior (second-level) distribution, \emph{i.e.}, a prior mean. Based on this conjugate prior distribution it is easy to derive the corresponding posterior distribution

\begin{equation} \label{normalpost}
\mu_{j}\vert \textbf{y}, \beta, A \stackrel{ind}{\sim}\textrm{Normal}(~(1-B_{j})y_{j} + B_{j}\mu_{0j},~(1-B_{j})V_{j}~),
\end{equation}
where $B_{j}\equiv V_{j}/(V_{j} + A),~j=1, \ldots, k$, are called shrinkage factors. By integrating over $\mu_{j}$ for $j = 1 \ldots k$ a marginal distribution, useful for model checking (Section \ref{discussion}), can be derived
\begin{equation} \label{normalmarginal}
y_{j}\vert \beta, A \stackrel{ind}{\sim}\textrm{Normal}(\mu_{0j}, V_{j} + A). 
\end{equation}

\subsection[Poisson-Gamma]{Poisson-Gamma (``p''=Poisson data)}\label{poissonsubsec}
\code{gbp} is also capable of estimating a Poisson-Gamma hierarchical model, though its usage is limited to the case where the second-level mean is known ($m=0$). For notational consistency, let's define $y_{j}\equiv z_{j} / n_{j}$ as the unbiased estimate of $\lambda_{j}$ for the $j$-th group, where  $z_{j}$ is the number of successes (or failures) out of $n_{j}$ trials. For this Poisson model $n_{j}$ can be interpreted as the exposure of the $j$-th group,  not necessarily an integer. The notation $r$ represents the amount of prior information as $n_{j}$ does the amount of observed information. Note that the constant, $1/r$, multiplied to the Gamma distribution below is a scaling factor and the square brackets indicates the [mean, variance] of a distribution. 
\begin{equation}
z_{j}\vert \lambda_{j} \stackrel{ind}{\sim}\textrm{Poisson}(n_{j}\lambda_{j}),
\end{equation}
\begin{equation}\label{gammaprior}
\lambda_{j}\vert \beta, r\stackrel{ind}{\sim}\frac{1}{r}\textrm{Gamma}(r\lambda_{0})\sim \textrm{Gamma} \left[\lambda_{0}, ~\frac{\lambda_{0}}{r} \right],
\end{equation}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 3,   2,   5,  11,   9,  12,  12,   4,  10,  13,  14,   7,  12,
          11,  13,  22,  15,  11,  14,  11,  16,  14,   9,  15,  13,  35,
          26,  25,  20,  35,  27)
R> n <- c(67,  68, 210, 256, 269, 274, 278, 295, 347, 349, 358, 396, 431,
         441, 477, 484, 494, 501, 505, 540, 563, 593, 602, 629, 636, 729,
         849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}
or
\begin{CodeChunk}
\begin{CodeInput}
R> data(hospital)
R> z <- hospital$d # variable name for the number of deaths is d in dataset
R> n <- hospital$n
\end{CodeInput}
\end{CodeChunk}

In addition, suppose one knows that the state-level death rate per exposure of this surgery is 0.030 ($\lambda_{0}$). Using these data and the known second-level mean ($\lambda_{0}$), \pkg{Rgbp} provides point and interval estimates of the true death rate ($\lambda_{j}$) so that one can evaluate each hospital's reliability.

@@

\subsection[Unknown Second-level Mean and No Covariate]{8 Schools: Unknown Second-level Mean with No Covariates} \label{sec:ex:8schools}

@@

Due to its speed and ease of use, \pkg{Rgbp} can be used for as a method of preliminary data analysis. Such results may tell statisticians and practitioners alike whether a more intensive method in terms of implementation and computational time, such as MCMC, is needed.
